---
title: "Backdown"
output: html_document
date: "2025-02-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Remarks : use Biostrings or seqinr packages to lighten code

```{r}
library(tidyverse)
library(forcats)  # Pour fct_reorder
```


# 1st step
Generate the RNA sequence with differents nitrogenous bases.
```{r}
gen_rna_seq <- function(n) {
  bases <- c("A", "U", "C", "G")
  paste0(sample(bases, n, replace = TRUE), collapse = "")
}
```

Split in groups of three elements (codons)
```{r}
split_into_codons <- function(sequence) {
  codons <- substring(sequence, seq(1, nchar(sequence)-2, by=3), seq(3, nchar(sequence), by=3))
  return(codons)
}
```

Library of AA corresponding to codons
```{r}
translate_codon <- function(codon) {
  genetic_code <- list(
    "UUU" = "Phenylalanine", "UUC" = "Phenylalanine", "UUA" = "Leucine", 
    "UUG" = "Leucine", "CUU" = "Leucine", "CUC" = "Leucine", "CUA" = "Leucine", 
    "CUG" = "Leucine", "AUU" = "Isoleucine", "AUC" = "Isoleucine", 
    "AUA" = "Isoleucine", "AUG" = "Methionine (Start)", "GUU" = "Valine", 
    "GUC" = "Valine", "GUA" = "Valine", "GUG" = "Valine", "UCU" = "Serine", 
    "UCC" = "Serine", "UCA" = "Serine", "UCG" = "Serine", "CCU" = "Proline", 
    "CCC" = "Proline", "CCA" = "Proline", "CCG" = "Proline", "ACU" = "Threonine",
    "ACC" = "Threonine", "ACA" = "Threonine", "ACG" = "Threonine", 
    "GCU" = "Alanine", "GCC" = "Alanine", "GCA" = "Alanine", "GCG" = "Alanine",
    "UAU" = "Tyrosine", "UAC" = "Tyrosine", "UAA" = "Stop", "UAG" = "Stop", 
    "UGA" = "Stop", "CAU" = "Histidine", "CAC" = "Histidine",
    "CAA" = "Glutamine", "CAG" = "Glutamine", "AAU" = "Asparagine", 
    "AAC" = "Asparagine", "AAA" = "Lysine", "AAG" = "Lysine",
    "GAU" = "Aspartic Acid", "GAC" = "Aspartic Acid",  "GAA" = "Glutamic Acid", 
    "GAG" = "Glutamic Acid", "UGU" = "Cysteine", "UGC" = "Cysteine",
    "UGG" = "Tryptophan", "CGU" = "Arginine", "CGC" = "Arginine", 
    "CGA" = "Arginine", "CGG" = "Arginine", "AGU" = "Serine", "AGC" = "Serine",
    "AGA" = "Arginine", "AGG" = "Arginine", "GGU" = "Glycine", "GGC" = "Glycine",
    "GGA" = "Glycine", "GGG" = "Glycine"
  )
  return(genetic_code[[codon]])
}
```

Amino Acids frequencies computing
```{r}
AA_frequencies <- function(sequence) {
  codons <- split_into_codons(sequence)
  amino_acids <- character(0)
  stop_codons <- 0  # count stop codons

  
  for (codon in codons) {
    aa <- translate_codon(codon)
    if (aa == "Stop") {
      stop_codons <- stop_codons + 1
    } else {
      amino_acids <- c(amino_acids, aa)
    }
  } 
  
  freq_table <- table(amino_acids)
  list(frequencies = freq_table, stop_codons = stop_codons)
}
```

# Test
```{r}
rna_sequence <- gen_rna_seq(1000)
result <- AA_frequencies(rna_sequence)
print(result$frequencies)
cat("Stop codons:", result$stop_codons, "\n")
```

# length of sequences and the repetition number of each
```{r}
sequence_lengths <- seq(100, 500, by=50)
n_repeats <- 100
data_list <- list()

for (len in sequence_lengths) {
  for (rep in 1:n_repeats) {
    rna_sequence <- gen_rna_seq(len)
    freq_data <- AA_frequencies(rna_sequence)
    
    df <- data.frame(
      Length = len,
      AminoAcid = names(freq_data$frequencies),
      Count = as.numeric(freq_data$frequencies)
    )
    data_list <- append(data_list, list(df))
  }
}
```

Compute proportions, reorder data for plot
```{r}
final_data <- bind_rows(data_list) %>%
  group_by(Length, AminoAcid) %>%
  summarise(MeanCount = mean(Count), .groups = "drop") %>%
  group_by(Length) %>%
  mutate(Proportion = MeanCount / sum(MeanCount)) %>%   # Standardization
  mutate(AminoAcid = forcats::fct_reorder(AminoAcid, Proportion, .desc = TRUE)) 
```



# Visualisation
```{r}
ggplot(final_data, aes(x = factor(Length), y = Proportion, fill = AminoAcid)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_viridis_d(option = "turbo") +  # Palette lisible et optimisée
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of Amino Acids by Sequence Length",
    x = "Sequence Length",
    y = "Amino Acids Proportion",
    fill = "Amino Acids"
  ) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```

# Table
```{r}
tableau_proportions <- final_data %>%
  select(Length, AminoAcid, Proportion) %>%
  tidyr::pivot_wider(names_from = AminoAcid, values_from = Proportion)
```

```{r}
print(tableau_proportions, n = Inf)
```

It seems that the result is implied by a correlation between the number of codons 
coding for each protein, hence the following verification, where we verifiy the 
monotonous positive relationship between these 2 variables:
```{r}
codon_count <- data.frame(
  AminoAcid = c("Phenylalanine", "Leucine", "Isoleucine", "Methionine (Start)", 
                "Valine", "Serine", "Proline", "Threonine", "Alanine", "Tyrosine",
                "Histidine", "Glutamine", "Asparagine", "Lysine", "Aspartic Acid", 
                "Glutamic Acid", "Cysteine", "Tryptophan", "Arginine", "Glycine"),
  CodonNumber = c(2, 6, 3, 1, 4, 6, 4, 4, 4, 2, 
                  2, 2, 2, 2, 2, 2, 2, 1, 6, 4)
)

final_data <- final_data %>%
  left_join(codon_count, by = "AminoAcid")

correlation_result <- cor(final_data$Proportion, final_data$CodonNumber, 
                          method = "spearman", use = "complete.obs")
print(paste("Spearman correlation:", round(correlation_result, 3)))

ggplot(final_data, aes(x = CodonNumber, y = Proportion)) +
  geom_point(aes(color = AminoAcid), size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  scale_color_viridis_d(option = "turbo") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Correlation between AA Proportion and Codon Redundancy",
    x = "Number of Codons per Amino Acid",
    y = "Mean Proportion",
    color = "Amino Acid"
  )

```

