---
title: "Backdown"
output: html_document
date: "2025-02-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Remarks : use Biostrings or seqinr packages to lighten code

```{r}
source("functions.R")
source("extradata.R")


library(tidyverse)
library(forcats)  # fct_reorder
library(tibble)
library(data.table)
```

# We define a seq length and a number of repetitions for each sequence value
```{r}
A <- Loop_AA_data(begin_size = 100, end_size = 500, step = 50, n_repeats = 10)
```

Compute proportions, reorder data for plot, and add melting temperature, that 
don't need to be added before neither in proportions nor in the repetitions
```{r}

```


```{r}
P <- ready2plot_data(A) %>%
  left_join(MeltingTemperatureData[c(2,4)], by ="AminoAcid")
  

print(P %>% group_by(Length) %>% slice_sample(n = 1))
```

# Visualisation
```{r}
ggplot(final_data, aes(x = factor(Length), y = Proportion, fill = AminoAcid)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_viridis_d(option = "turbo") +  
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of Amino Acids by Sequence Length",
    x = "Sequence Length",
    y = "Amino Acids Proportion",
    fill = "Amino Acids"
  ) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )




# Plot computed temperature with AA table
ggplot(final_data, aes(x= MeltTemp, y=Proportion)) + 
  geom_point(aes(color = AminoAcid), size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +  
  scale_fill_viridis_d(option = "turbo") +  
  theme_minimal(base_size = 14) +
  labs(
    title = "Distribution of Amino Acids by melting temperature",
    x = "Melting Temperature (°C)",
    y = "Amino Acids Proportion",
    fill = "Amino Acids"
  ) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```


# Complementary Strand
```{r}
# Generate complementary strand and compute frequencies
rna_complementary_seq <- gen_comp_rna_seq(rna_sequence)
result_comp <- AA_frequencies(rna_complementary_seq)

# Store complementary strand data
data_list_comp <- list()
for (len in sequence_lengths) {
  for (rep in 1:n_repeats) {
    rna_complementary_seq <- gen_comp_rna_seq(gen_rna_seq(len))
    freq_data_comp <- AA_frequencies(rna_complementary_seq)

    df_comp <- data.frame(
      Length = len,
      AminoAcid = names(freq_data_comp$frequencies),
      Count = as.numeric(freq_data_comp$frequencies)
    )
    data_list_comp <- append(data_list_comp, list(df_comp))
  }
}

# Process complementary strand data
final_data_comp <- bind_rows(data_list_comp) %>%
  group_by(Length, AminoAcid) %>%
  summarise(MeanCount = mean(Count), .groups = "drop") %>%
  group_by(Length) %>%
  mutate(Proportion = MeanCount / sum(MeanCount) * length(sequence_lengths)) %>%
  mutate(AminoAcid = forcats::fct_reorder(AminoAcid, Proportion, .desc = TRUE)) %>%
  left_join(data.frame(AminoAcid = names(result_comp$tm_values), MeltTemp = result_comp$tm_values), by = "AminoAcid")


# Plot complementary strand amino acid proportions
ggplot(final_data_comp, aes(x = factor(Length), y = Proportion, fill = AminoAcid)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_viridis_d(option = "turbo") +
  theme_minimal(base_size = 14) +
  labs(
    title = "(Complementary Strand)",
    x = "Sequence Length",
    y = "Amino Acids Proportion",
    fill = "Amino Acids"
  ) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )



# Plot complementary strand melting temperature vs proportion
ggplot(final_data_comp, aes(x= MeltTemp, y=Proportion)) + 
  geom_point(aes(color = AminoAcid), size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +  
  scale_fill_viridis_d(option = "turbo") +  
  theme_minimal(base_size = 14) +
  labs(
    title = "(Complementary Strand)",
    x = "Melting Temperature",
    y = "Amino Acids Proportion",
    fill = "Amino Acids"
  ) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

#ggsave("Cstrand_AA_distribution_MeltTemp_RandomDist.jpeg", width = 8, height = 6, dpi = 300)

```


## Comparison
```{r}
# Define the number of repetitions
num_repetitions <- 100

# Store the results of each repetition in a list
results_list <- list()

for (i in 1:num_repetitions) {
  # Generate random RNA sequence
  rna_sequence <- gen_rna_seq(500)  # Example RNA sequence
  comp_rna_sequence <- gen_comp_rna_seq(rna_sequence)
  
  # Get amino acid frequencies for both sequences
  result_original <- get_amino_acid_frequencies(rna_sequence)
  result_complementary <- get_amino_acid_frequencies(comp_rna_sequence)
  
  # Ensure both result tables are using the same set of amino acids
  all_aminos <- unique(c(names(result_original$frequencies), names(result_complementary$frequencies)))
  
  # Initialize the frequencies (to ensure both sequences have frequencies for all amino acids)
  original_freq <- rep(0, length(all_aminos))
  complementary_freq <- rep(0, length(all_aminos))
  
  # Set the names for easy lookup
  names(original_freq) <- all_aminos
  names(complementary_freq) <- all_aminos
  
  # Populate the frequencies from the results
  original_freq[names(result_original$frequencies)] <- result_original$frequencies
  complementary_freq[names(result_complementary$frequencies)] <- result_complementary$frequencies
  
  # Create a data frame with amino acid frequencies for both sequences
  comparison_data <- data.frame(
    AminoAcid = all_aminos,
    Original_Proportion = original_freq / sum(original_freq),
    Complementary_Proportion = complementary_freq / sum(complementary_freq)
  )
  
  # Reorder AminoAcid based on the absolute difference in proportions
  comparison_data <- comparison_data %>%
    mutate(AminoAcid = fct_reorder(AminoAcid, abs(Original_Proportion - Complementary_Proportion), .desc = TRUE))
  
  # Append the result to the list
  results_list[[i]] <- comparison_data
}


# Combine all the individual results into one large data frame
long_term_data <- bind_rows(results_list)

# Calculate the mean differences across all repetitions
mean_comparison_data <- long_term_data %>%
  group_by(AminoAcid) %>%
  summarise(
    Mean_Original_Proportion = mean(Original_Proportion),
    Mean_Complementary_Proportion = mean(Complementary_Proportion),
    Mean_Proportion_Difference = mean(abs(Original_Proportion - Complementary_Proportion)),
    .groups = 'drop'
  ) %>%
  mutate(AminoAcid = fct_reorder(AminoAcid, Mean_Proportion_Difference, .desc = TRUE))

# Plot the mean differences
ggplot(mean_comparison_data, aes(y = AminoAcid, x = Mean_Proportion_Difference)) +
  geom_bar(stat = "identity", fill = "gray", width = 0.6) +
  labs(title = "Average Differences in Amino Acid Proportions: Original vs Complementary",
       x = "Mean Proportion Difference (Original - Complementary)", y = "Amino Acid") +
  scale_x_continuous() +
  theme_minimal()

ggsave("Comp_Original_strands_proportion_comparison.jpeg", width = 8, height = 6, dpi = 300)



```

### Table
```{r}
tableau_proportions <- final_data %>%
  select(Length, AminoAcid, Proportion) %>%
  tidyr::pivot_wider(names_from = AminoAcid, values_from = Proportion)
```

```{r}
print(tableau_proportions, n = Inf)
```

It seems that the result is implied by a correlation between the number of codons 
coding for each protein and the proportion of those, hence the following 
verification, where we verifiy the monotonous positive relationship between 
these 2 variables:
```{r}
codon_count <- data.frame(
  AminoAcid = c("Phenylalanine", "Leucine", "Isoleucine", "Methionine (Start)", 
                "Valine", "Serine", "Proline", "Threonine", "Alanine", "Tyrosine",
                "Histidine", "Glutamine", "Asparagine", "Lysine", "Aspartic Acid", 
                "Glutamic Acid", "Cysteine", "Tryptophan", "Arginine", "Glycine"),
  CodonNumber = c(2, 6, 3, 1, 4, 6, 4, 4, 4, 2, 
                  2, 2, 2, 2, 2, 2, 2, 1, 6, 4)
)

final_data <- final_data %>%
  left_join(codon_count, by = "AminoAcid")

correlation_result <- cor(final_data$Proportion, final_data$CodonNumber, 
                          method = "spearman", use = "complete.obs")
print(paste("Spearman correlation:", round(correlation_result, 3)))

ggplot(final_data, aes(x = CodonNumber, y = Proportion)) +
  geom_point(aes(color = AminoAcid), size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") +
  scale_color_viridis_d(option = "turbo") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Correlation between AA Proportion and Codon Redundancy",
    x = "Number of Codons per Amino Acid",
    y = "Mean Proportion",
    color = "Amino Acid"
  )

```









